Physcraper: Body

### 2020-08-19 2-body.Rmd
second draft final :tada:

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local database sequence searches.
  For remote database searches, we modified the BioPython [@cock2009biopython] BLAST function from the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the "physcraper" folder, allowing for secondary analyses to run significantly faster.
BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80% and 120%, respectively).
  This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
  The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
All sequences accepted up to this point are assigned an internal identifier.
New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.
Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
All BLAST and filtering parameters can be customized by the user.
Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions [@cock2009biopython].
Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
New sequences passing all filtering steps are added to the "csv" taxon summary file.
A "fasta" file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

By default, Physcraper uses the software MUSCLE [@edgar2004muscle] to perform DNA sequence alignments. Instructions on how to install all software dependencies used by Physcraper are provided in the documentation.
The process to align new sequences consists of two steps. First, all new sequences are aligned using the default MUSCLE options.
Second, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align the new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing [e.g., GBlocks; @castresana2000selection; @castresana2002gblocks].
While curating the alignment is a critical step, it is not a reproducible one. The main reason for its lack of reproducibility might be that it is hard to track changes made on the alignment. A form of version control, to register the differences between the alignment that was produced by the software and the manually curated alignment would ideal.
Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.

@@

A Maximum Likelihood (ML) gene tree is reconstructed for each alignment provided, using the software RAxML [@stamatakis2014raxml] with default settings, such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with the default algorithm. Currently only the number of bootsrap replicates can be specified by the user.
By default, the original tree is used as a starting tree for the ML searches. Alternatively, users can set the original tree as a full topological constraint, or ignore it completely for the searches.
Bootstrap results are summarized with the SumTrees module of DendroPy [current version 4.4.0; @sukumaran2010dendropy].
Physcraper's final result is an updated phylogenetic hypothesis for the locus provided in the input alignment.
Tips on all trees generated by Physcraper are defined by a taxon "name space". The taxon metadata captures the NCBI accession information, as well as the taxon identifiers, allowing the user to perform comparisons and conflict analyses.
Two ways to compare the updated tree with the original tree are implemented in Physcraper. First, Robinson Foulds weighted and unweighted metrics are estimated using Dendropy functions [@sukumaran2010dendropy].
Second, a conflict analysis is performed. This is a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined. A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroup directly on the updated tree, so trees are accurately rooted.
<!-- *It would be a nice addition to have users give the output of the input tree as an argument at some point, or maybe we could add a super outgroup at random based on the search taxon* -->

@@

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The best RaxML tree is 99% resolved, while the consensus tree is 75% resolved, meaning that 25% of nodes have a bootstrap < 0.1, while 48% nodes have bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches < 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75 (Fig. \@ref(fig:results)).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in the updated Gottlieb2005 tree. @yao2020phylogeny also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis.
When those sequences are available online, repeating the physcraper run will easily incorporate them into the expanded analysis.
The Yao2020 tree also lacks 18 taxa that were added by Physcraper.
<!-- **Any statements about relationships or anything? Are there multiple seqs for these taxa?** -->
This might be caused by the method they used to download existing ITS *Ilex* sequences from GenBank, which is not fully explained in the publication.
  Other examples of Physcraper runs on empirical data at a range of taxonomic scales are available in the Physcraper documentation online.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE ?-->

>>>>

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.

The Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped] is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The `blastn` function from the BLAST command line tools [@camacho2009blast] is used for local database sequence searches.
  For remote database searches, we modified the BioPython [@cock2009biopython] BLAST function from the [NCBIWWW module](https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html) to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the "physcraper" folder, allowing for secondary analyses to run significantly faster.

BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80% and 120%, respectively).
  This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
  The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
All sequences accepted up to this point are assigned an internal identifier.
New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.
Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.

All BLAST and filtering parameters can be customized by the user.
Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions [@cock2009biopython].
Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
New sequences passing all filtering steps are added to the "csv" taxon summary file.
A "fasta" file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

By default, Physcraper uses the software MUSCLE [@edgar2004muscle] to perform DNA sequence alignments. Instructions on how to install all software dependencies used by Physcraper are provided in the documentation.
The process to align new sequences consists of two steps. First, all new sequences are aligned using the default MUSCLE options.

Second, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align the new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing [e.g., GBlocks; @castresana2000selection; @castresana2002gblocks].
While curating the alignment is a critical step, it is not a reproducible one. The main reason for its lack of reproducibility might be that it is hard to track changes made on the alignment. A form of version control, to register the differences between the alignment that was produced by the software and the manually curated alignment would ideal.
Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.

@@

A Maximum Likelihood (ML) gene tree is reconstructed for each alignment provided, using the software RAxML [@stamatakis2014raxml] with default settings, such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with the default algorithm. Currently only the number of bootsrap replicates can be specified by the user.
By default, the original tree is used as a starting tree for the ML searches. Alternatively, users can set the original tree as a full topological constraint, or ignore it completely for the searches.
Bootstrap results are summarized with the SumTrees module of DendroPy [current version 4.4.0; @sukumaran2010dendropy].

Physcraper's final result is an updated phylogenetic hypothesis for the locus provided in the input alignment.
Tips on all trees generated by Physcraper are defined by a taxon "name space". The taxon metadata captures the NCBI accession information, as well as the taxon identifiers, allowing the user to perform comparisons and conflict analyses.
Two ways to compare the updated tree with the original tree are implemented in Physcraper. First, Robinson Foulds weighted and unweighted metrics are estimated using Dendropy functions [@sukumaran2010dendropy].
Second, a conflict analysis is performed. This is a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface [@redelings2017supertree].
For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined. A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroup directly on the updated tree, so trees are accurately rooted.
<!-- *It would be a nice addition to have users give the output of the input tree as an argument at some point, or maybe we could add a super outgroup at random based on the search taxon* -->

@@

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The best RaxML tree is 99% resolved, while the consensus tree is 75% resolved, meaning that 25% of nodes have a bootstrap < 0.1, while 48% nodes have bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches < 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75 (Fig. \@ref(fig:results)).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in the updated Gottlieb2005 tree. @yao2020phylogeny also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis.
When those sequences are available online, repeating the physcraper run will easily incorporate them into the expanded analysis.
The Yao2020 tree also lacks 18 taxa that were added by Physcraper.
<!-- **Any statements about relationships or anything? Are there multiple seqs for these taxa?**
add 1 sentence saying that it is odd that we are getting differnet relationships given what is likely to be the same data
but we can't compare the exact seqs or alignemnt, bc they are not available.
 -->
This might be caused by the method they used to download existing ITS *Ilex* sequences from GenBank, which is not fully explained in the publication.
  Other examples of Physcraper runs on empirical data at a range of taxonomic scales are available in the Physcraper documentation online.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE ?-->
